---
title: "Continuous Characters"
output:
  html_document: default
---

```{r, cache=TRUE}
library(ape)
library(geiger)
library(OUwie)
library(gdata)
library(RCurl)

tree<-read.tree(text=getURL("https://raw.githubusercontent.com/lukejharmon/pcm/master/datafiles/squamate.phy"))
sqdata <- read.csv(text=getURL("https://raw.githubusercontent.com/lukejharmon/pcm/master/datafiles/brandley_table.csv"), row.names = 1, check.names = TRUE, stringsAsFactors = FALSE)
rownames(sqdata) <- sub(" ", "_", rownames(sqdata))

#putting the continuous data in one object
contin_data <- as.matrix(sqdata[,2])
rownames(contin_data) <- row.names(sqdata)
class(contin_data)

discrete_data <- sqdata[,c(15, 16)]

#Clean data
CleanData <- function(phy, data) {
  treedata(phy = phy, data = data, sort = TRUE, warnings = TRUE)
}

cleaned_data <- CleanData(tree, contin_data)
cleaned_discrete <- CleanData(phy = tree, data = discrete_data)

#Visualizing data
cleaned_tree <- cleaned_data$phy
chrono_tree <- ape::chronos(cleaned_tree)
ape::is.ultrametric(chrono_tree)
cleaned_contin <- as.vector(cleaned_data$data)
is.vector(cleaned_contin)
names(cleaned_contin) <- make.names(row.names(cleaned_data$data))

VisualizeData <- function(phy, data) {
 pretty_pic <- phytools::contMap(tree = phy, x = data, fsize = c(.2,1), lwd = 2)
 polytomy <- is.binary(phy = phy)
 pretty_pic
 print(paste("Is the tree binary?", polytomy))
}

str(cleaned_contin)

VisualizeData(chrono_tree, cleaned_contin)

#Rate of Evolution
BM1 <- geiger::fitContinuous(chrono_tree, cleaned_contin, model="BM")
print(paste("The rate of evolution is", BM1$opt$sigsq, "in units of mm/million years" ))

OU1 <- fitContinuous(chrono_tree, cleaned_contin, model="OU")
par(mfcol=c(1,2))
plot(chrono_tree, show.tip.label=FALSE)
ou.tree <- geiger::rescale(chrono_tree, model="OU", alpha = OU1$opt$alpha)
plot(ou.tree, cex = .2)

```
  
The trees differ in their branch lengths. The rescaled tree condenses the branch lengths of older part of the tree and expands it closer to the tips.  

```{r, cache=TRUE}
AIC_BM1 <- BM1$opt$aic
AIC_OU1 <- OU1$opt$aic
delta_AIC_BM1 <- BM1$opt$aic - BM1$opt$aicc
delta_AIC_OU1 <- OU1$opt$aic - OU1$opt$aicc
all_aic <- matrix(c(AIC_BM1, delta_AIC_BM1, AIC_OU1, delta_AIC_OU1),ncol = 1)
rownames(all_aic) <- c("BM1 AIC", "delta BM1 AIC", "OU1 AIC", "delta OU1 AIC")
all_aic <- as.table(all_aic)
print(all_aic)

one_discrete_char <- cleaned_discrete$dat[,1]
reconstruction.info <- ape::ace(one_discrete_char, cleaned_discrete$phy, type="discrete", method="ML", CI=TRUE)
best.states <- colnames(reconstruction.info$lik.anc)[apply(reconstruction.info$lik.anc, 1, which.max)]

chrono_tree$node.labels <- best.states

clean_data_df <- as.data.frame(cleaned_data$data)
library(data.table)
library(tibble)
clean_data_df <- data.table(clean_data_df, keep.rownames = TRUE)
clean_data_df <- add_column(clean_data_df, one_discrete_char, .after = 1)

nodeBased.OUMV <- OUwie::OUwie(chrono_tree, clean_data_df, model="OUMV", simmap.tree=FALSE, diagn=FALSE)
print(nodeBased.OUMV)
```

```{r, cache=TRUE}
models <- c("BM1","BMS","OU1","OUM","OUMV","OUMA","OUMVA")
results <- lapply(models, OUwie::OUwie, phy=chrono_tree, data=clean_data_df)

AICc.values<-sapply(results, "[[", "AICc")
names(AICc.values)<-models
AICc.values<-AICc.values-min(AICc.values)


print(AICc.values) #The best model is the one with smallest AICc score

best<-results[[which.min(AICc.values)]] #store for later

print(best) #prints info on best model

#I picked alpha values that tried to cover all of the alpha values for the different states from the best model.
alpha.values<-seq(from= .1, to= 5, length.out=50)

likelihood.values <- rep(NA, length(alpha.values))
for (iteration in sequence(length(alpha.values))) {
    likelihood.values[iteration] <- OUwie::OUwie.fixed(chrono_tree, clean_data_df, model="OUMV", alpha=rep(alpha.values[iteration], times=1), sigma.sq=best$solution[2,], theta=best$theta[,1])$loglik[1]
}
```
```{r, cache=TRUE}
plot(x= alpha.values, y= likelihood.values, xlab="Alpha", ylab="Likelihood", type="l", bty="n", ylim = c(-1580, -1370))
points(x=best$solution[1,1], y=best$loglik, pch=16, col="red")
text(x=best$solution[1,1], y=best$loglik, "unconstrained best", pos=4, col="red")
abline(h=best$loglik-2, lty="dotted")

library(akima)
nreps<-400
theta1.points<-c(best$theta[1,1], rnorm(nreps-1, best$theta[1,1], 5*best$theta[1,2])) #center on optimal value, have extra variance
theta2.points<-c(best$theta[2,1], rnorm(nreps-1, best$theta[2,1], 5*best$theta[2,2])) #center on optimal value, have extra variance
theta3.points<-c(best$theta[3,1], rnorm(nreps-1, best$theta[3,1], 5*best$theta[3,2]))

likelihood.values<-rep(NA,nreps)

for (iteration in sequence(nreps)) {
    likelihood.values[iteration] <- OUwie::OUwie.fixed(chrono_tree, clean_data_df, model="OUMV", alpha=best$solution[1,], sigma.sq=best$solution[2,], theta=c(theta1.points[iteration], theta2.points[iteration], theta3.points[iteration]))$loglik
}

likelihood.differences<-(-(likelihood.values-max(likelihood.values)))

```
```{r}
head(theta1.points)
head(theta2.points)
head(theta3.points)
```
My discrete data had 3 states (0,1,and 2), so I had to add an additional theta vector to the iteration. I'm not sure how to add an additional dimension to get those all to plot against the likelihood differences. Also, the range between theta 1 and 2 are too different to be able to plot.  

```{r}
interpolated.points<-akima::interp(x=theta1.points, y=theta3.points, z= likelihood.differences, linear=FALSE, extrap=TRUE, xo=seq(min(theta1.points), max(theta1.points), length = 400), yo=seq(min(theta3.points), max(theta3.points), length = 400))

contour(interpolated.points, xlim=range(c(theta1.points, theta3.points)),ylim=range(c(theta1.points, theta3.points)), xlab="Theta 1", ylab="Theta 3", levels=c(2,5,10),add=FALSE,lwd=1, bty="n", asp=1)

points(x=best$theta[1,1], y=best$theta[3,1], col="red", pch=16)

points(x=clean_data_df$one_discrete_char[which(clean_data_df$Reg==0)],y=rep(min(c(theta1.points, theta3.points)), length(which(clean_data_df$Reg==0))), pch=18, col=rgb(0,0,0,.3)) #the tip values in regime 1, plotted along x axis
points(y=clean_data_df$one_discrete_char[which(clean_data_df$Reg==2)],x=rep(min(c(theta1.points, theta3.points)), length(which(clean_data_df$Reg==2))), pch=18, col=rgb(0,0,0,.3)) #the tip values in regime 2, plotted along y axis

library(phytools)
trait.ordered<-data.frame(clean_data_df[,2], clean_data_df[,2], row.names=clean_data_df[,1])
trait.ordered<- trait.ordered[chrono_tree$tip.label,]
z<-trait.ordered[,1]
names(z)<-rownames(trait.ordered)
tree.mapped<-make.simmap(chrono_tree,z,model="ER",nsim=1)
leg<-c("black","red")
names(leg)<-c(1,2)
plotSimmap(tree.mapped,leg,pts=FALSE,ftype="off", lwd=1)

simmapBased<-OUwie::OUwie(tree.mapped,clean_data_df,model="OUMV", simmap.tree=TRUE, diagn=FALSE)
print(simmapBased)
print(best)
```
Based on the AIC values, the best model from above was better than the simmap based one. These two probably shouldn't be directly comparable, at least with this dataset since the sigsq values are so high for the 0 and 1 states.  